Powering Hidden Markov Model by Neural Network based Generative Models

### 2019-08-20 manuscript/main.tex
draft v1

\title{Powering Hidden Markov Model by Generative Models}
\author{
  Dong Liu, Antoine Honoré, Saikat Chatterjee, and Lars~K. Rasmussen\\
  KTH Royal Institute of Technology, Stockholm, Sweden \\
  E-mail: \{doli, honore, sach, lkra\}@kth.se}

@@

\maketitle
\begin{abstract}
  to be written...
\end{abstract}

\section{introduction}
Sequential data modeling is a challenging topic in pattern recognition and machine learning, since for many applications, the assumption of independent and identically distributed data points is too strong to model data properly. Hidden Markov model (HMM) is classic way to model sequential data. HMM has been widely used in different practical problems, and especially is known for applications in reinforcement learning \cite{ding2018reinforcementhmm,levine2018reinforcementReview}, natural language modeling \cite{khan2016survey,Hariyanti_2019}, biological sequence analysis such as proteins \cite{ASHWIN20172} and DNA \cite{ren2015dna}, etc.

HMM provides an framework for sequential data modeling. To connect states of HMM with input data , a probabilistic model is used to describe how each state of HMM should represent a part of sequential data. The typical way is to use a Gaussian mixture model (GMM) per state of HMM\cite{juang1986maximum}, where GMMs are used to represent the relationship between states of HMM and sequential data input. GMM based HMM (GMM-HMM) has becomes a standard model for sequential data modeling, and employed widely for practical applications, especially in speech recognition \cite{gales2008application}.

Given the success of GMM-HMM, it is not efficient for modeling data in nonlinear manifold. Research attempts of training HMM with neural networks have been done to boost the modeling capacity of HMM. A successful work is to bring restrictive Boltzmann machine (RBM) \cite{Hinton2012} into HMM learning \cite{hinton2012deepSpeech,li2013hybrid,Miao2013ImprovingLC}. In RBM based HMM, the hierarchical learning scheme is used: i) a stack of RBMs are trained in unsupervised fashion, ii) then a final softmax layer is added to the stack of RBMs to represent the posterior of state of HMM given data input, iii) a discriminative training is followed to do the final tuning of the model.

@@

work of epfl fiting GMM with neural network...

The above work help improve modeling capacity of HMM based models by bringing in neural networks. A softmax layer is usually used to represent probability whenever a conditional distribution is needed. These hierarchical schemes are built based on intuition of domain knowledge. Training of these hierarchical models usually need specific experience in relevant area to be able to the hierarchical procedure of training.

In this work, we propose a generator-mixed HMM model, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
\item has advantage of easy to do training, since GenHMM training employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, but configure of GenHMM is flexible.
\item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM is able to compute exact loglikelihood of given sequence data based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. The likelihood computation of sequential data under GenHMM can be computed exactly and efficiently.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Convergence analysis for GenHMM with the learning algorithm with gradient based optimization.
\end{itemize}

@@

Our framework is an Hidden Markov Model (HMM). An HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ means transpose, and $T$ denote the time length. Using HMM for signal representation or generating is illustrated in \ref{fig:hmm}. The assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by different source, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of states of HMM $\bm{H}$;
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ initial distribution of HMM $\bm{H}$ with $|\Ss|$ is cardinality of $\Ss$, $q_k = p(s=k)$ for random state variable $s$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$, $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i)$.
\item For given hidden state $s$, the observable signal density is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this conditional probabilistic model.
\end{itemize}

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is generated by density $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This process gives us the probabilistic model $p(\ubm{x};\bm{H})$
\subsection{Emission Model of GenHMM}

@@

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as typical HMM, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For given state $s\in \Ss$, we define its probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, and $\pi_{s, \kappa} = p(\kappa | s; \bm{\Phi}_{s, \kappa})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]$,
$\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. 
We define $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}\sim p_{s,\kappa}(\bm{z})$. By change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

@@

For a given dataset, we denote its empirical distribution by $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$, where superscipt $(\cdot)^{r}$ denotes the index of sequential signal. The natural question is how to use GenHMM to represent the dataset. Alternative, we are looking for the answer to question:
\begin{equation}
  \umin{\bm{H}\in \Hh} KL(\hat{p}(\ubm{x})\| p(\ubm{x};\bm{H}))
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence. The KL divergence problem can be boiled down to a maximizing a loglikelihood problem:
\begin{equation}\label{eq:ml-of-hmm}
  \uargmax{\bm{H} \in \Hh} \sum_{r=1}^{R}\log\,p(\ubmr{x}{r}; \bm{H}).
\end{equation}

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$ and $\ubm{\kappa}$, we can not directly solve the maximum likelihood problem in \ref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H})\right],
  \end{equation}
  where $\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \cdot\right]$ denotes the expectation operator by distribution $\hat{p}(\ubm{x})$ and $p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})$.
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

The \eqref{eq:em-m-opt} can be reformulated as:
\begin{align}\label{eq:m-step-subs}
  &\umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =&\umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{{A}}\Qq({A}; \bm{H}^{\mathrm{old}}) 
     + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{align}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) 
  &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right] \nonumber\\
  &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right]\label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \sum_{t}\log\,p({s}_{t+1}|{s}_{t}; \bm{H}) \right] \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p(\ubm{x}, \ubm{\kappa}| \ubm{s}; \bm{H}) \right]\label{eq:generative-model-update}
\end{align}

We can see that the solution of $H$ depends on the posterior probability $p(\ubm{s}| \ubm{x}; \bm{H})$. Though the evaluation of posterior according to Bayesian theorem is simple, the computation complexity of $p(\ubm{s}| \ubm{x}; \bm{H})$ grows exponentially with the length of $\ubm{s}$. Therefore, we would employ Forward/Backward algorithm \cite{} to do the posterior computation efficiently. The marginal $p(s_t| \ubm{x}; \bm{H})$ is also efficiently computed as the joint posterior.

With such a model GenHMM at hand, the remaining problems are:
\begin{itemize}
\item How to realize GenHMM by neural networks to have high modeling capacity?
\item How to train GenHMM to sovle problem in \eqref{eq:ml-of-hmm} using practical algorithm?
\item Would the training of GenHMM converges?
\end{itemize}
We would answer these questions in the following section.

@@

In this section, we provide answers to the questions at the end of last section.

\subsection{Modeling $\bm{g}_{s,\kappa}$ by a flow model }
We use feed-forward neural network to implement every generator $\bm{g}_{s,\kappa}$. For notation simplicity, denote $\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}$. Define  a feed-forward neural network $\bm{g}_{s,\kappa}$ that has multiple hidden layers
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$ and is invertible $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$, where $\circ$ denotes mapping concatenation. Then the signal flow can be depicted as
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{x}} = \bm{h}_L$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture},
\end{equation*}
where $\bm{g}_{s,\kappa}^{[l]}$ and $\bm{f}_{s,\kappa}^{[l]}$ are the $l$-th layer of $\bm{g}_{s,\kappa}$ and $\bm{f}_{s,\kappa}$, respectively. In a feed-forward neural network, if every layer is invertible,
the full feed-forward neural network is invertible. The inverse
function is given by $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. Flow-based
network, proposed in \cite{DBLP:journals/corr/DinhKB14}, is such a
feed-forward neural network, which is further improved in subsequent works \cite{2016arXiv160508803D,kingma2018glow}. It also has additional advantages as efficient Jacobian computation and low computational complexity. %Computation of inverse mapping for flow-based network and determinant of Jacobin were addressed in the literature.

>>>>

\title{Powering Hidden Markov Model by Neural Network based Generative Models}
\author{
  Dong Liu$^{\dagger}$, Antoine Honoré$^{\ast \dagger}$, Saikat Chatterjee$^{\dagger}$, and Lars~K. Rasmussen$^{\dagger}$\\
  $^{\dagger}$KTH Royal Institute of Technology, Stockholm, Sweden \\
  $^{\ast}$Karolinska Institute, Stockholm, Sweden \\
  E-mail: \{doli, honore, sach, lkra\}@kth.se}

@@

\maketitle
\begin{abstract}
  Hidden Markov model (HMM) has been successfully used for sequential data modeling problems. In this work, we propose to power the modeling capacity of HMM by bringing in neural network based generative models. The proposed model is termed as GenHMM. In GenHMM, the hidden states of HMM and input data are related by neural network based generative model that provides exact likelihood computation. This is realized by a flow model. Apart from modeling itself, we also give learning algorithm for GenHMM in expectation-maximization framework. The convergence of learning GenHMM is analyzed. We demonstrate the efficiency of GenHMM by classification tasks on practical sequential data.
\end{abstract}

\section{introduction}
Sequential data modeling is a challenging topic in pattern recognition and machine learning, since for many applications, the assumption of independent and identically distributed data points is too strong to model data properly. Hidden Markov model (HMM) is a classic way to model sequential data. HMM has been widely used in different practical problems, and especially is known for applications in reinforcement learning \cite{ding2018reinforcementhmm,levine2018reinforcementReview}, natural language modeling \cite{khan2016survey,Hariyanti_2019}, biological sequence analysis such as proteins \cite{ASHWIN20172} and DNA \cite{ren2015dna}, etc.

HMM provides an framework for sequential data modeling. To connect states of HMM with input data, a probabilistic model is used to describe how each state of HMM should represent a pattern of sequential data. The typical way is to use a Gaussian mixture model (GMM) per state of HMM \cite{juang1986maximum}, where GMMs are used to represent the relationship between states of HMM and sequential data input. GMM based HMM (GMM-HMM) has becomes a standard model for sequential data modeling, and employed widely for practical applications, especially in speech recognition \cite{gales2008application,chatterjee2011auditory}.

Given the success of GMM-HMM, it is not efficient for modeling data in nonlinear manifold. Research attempts of training HMM with neural networks have been made to boost the modeling capacity of HMM. A successful work is to bring restrictive Boltzmann machine (RBM) \cite{Hinton2012} into HMM based models \cite{hinton2012deepSpeech,li2013hybrid,Miao2013ImprovingLC}. In RBM based HMM, a hierarchical learning scheme is used: i) a stack of RBMs are trained one after another in unsupervised fashion, ii) then a final softmax layer is added to the stack of RBMs to represent the posterior of state of HMM given data input, iii) a discriminative training is followed to do the final tuning of the model.

@@

% work of epfl fiting GMM with neural network...

The above work help improve modeling capacity of HMM based models by bringing in neural networks. A softmax layer is usually used to represent probability whenever a conditional distribution is needed. These hierarchical schemes are built based on intuition of domain knowledge. Training of these hierarchical models usually need specific experience in relevant area to be able to proceed with the hierarchical procedure of training.

In this work, we propose a generator-mixed HMM model, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
  \item firstly has high modeling capacity of sequential data, due to the neural network based generators;
\item has advantage of easy to do training, since GenHMM training employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible.
% \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM is able to compute exact loglikelihood of given sequence data based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. The likelihood computation of sequential data under GenHMM can be computed exactly and efficiently.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Convergence analysis for GenHMM under the proposed learning algorithm.
\end{itemize}

@@

Our framework is a Hidden Markov Model (HMM). A HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ means transpose, and $T$ denote the time length. Using HMM for signal representation or generating is illustrated in \ref{fig:hmm}. The assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by different source, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of states of HMM $\bm{H}$;
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ initial distribution of HMM $\bm{H}$ with $|\Ss|$ is cardinality of $\Ss$, $q_s = p(s;\bm{H})$ for random state variable $s$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$, $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For given hidden state $s$, the observable signal density function is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this conditional probabilistic model.
\end{itemize}

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is generated by density $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This process gives us the probabilistic model $p(\ubm{x};\bm{H})$.
\subsection{Generative Model of GenHMM}

@@

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as typical HMM, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For given state $s\in \Ss$, we define its probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, and $\pi_{s, \kappa} = p(\kappa | s; \bm{H})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]$,
$\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. 
We define $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}$ is a latent variable following a distribution with density function $p_{s,\kappa}(\bm{z})$. By change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

@@

For a given dataset, we denote its empirical distribution by $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$, where superscipt $(\cdot)^{r}$ denotes the index of sequential signal. The natural question is how to use GenHMM to represent the dataset. Alternative, we are looking for the answer to question:
\begin{equation}
  \umin{\bm{H}\in \Hh} KL(\hat{p}(\ubm{x})\| p(\ubm{x};\bm{H})),
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence. The KL divergence problem can be boiled down to a maximizing a loglikelihood problem:
\begin{equation}\label{eq:ml-of-hmm}
  \uargmax{\bm{H} \in \Hh} \sum_{r=1}^{R}\log\,p(\ubmr{x}{r}; \bm{H}).
\end{equation}

The first problem that we need to address is to deal with the hidden sequential variable $\ubm{s}$ and $\ubm{\kappa}$  in model $\bm{H}$. For a sequential observable variable $\ubm{x}$, $\ubm{s}$ is the hidden state sequence corresponding to $\ubm{x}$, and $\ubm{\kappa}$ is the hidden variable sequence representing the generator identity sequence that actually generates $\ubm{x}$.

Since directly maximizing likelihood is not an option for our problem in \ref{eq:ml-of-hmm}, we address this problem in expectation maximization (EM) framework. This divides our problem into two steps: i) using the joint posterior of hidden variable sequences $\ubm{s}$ and $\ubm{\kappa}$ to obtain a ``expected likelihood'' of observable variable sequence $\ubm{x}$, i.e. the E-step; ii) maximizing the expected likelihood with regarding to (w.r.t.) the model $\bm{H}$, i.e. the M-step. We formulate these two steps as follows:
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  the expected likelihood function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H})\right],
  \end{equation}
  where $\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \cdot\right]$ denotes the expectation operator by distribution $\hat{p}(\ubm{x})$ and $p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})$.
\item M-step: the maximization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}).
  \end{equation}
\end{itemize}

The problem \eqref{eq:em-m-opt} can be reformulated as:
\begin{align}\label{eq:m-step-subs}
  &\umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =&\umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}}) 
     + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{align}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) 
  % &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right] \nonumber\\
  &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1};\bm{H})  \right], \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \sum_{t}\log\,p({s}_{t+1}|{s}_{t}; \bm{H}) \right], \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p(\ubm{x}, \ubm{\kappa}| \ubm{s}; \bm{H}) \right]. \label{eq:generative-model-update}
\end{align}

We can see that the solution of $\bm{H}$ depends on the posterior probability $p(\ubm{s}| \ubm{x}; \bm{H})$. Though the evaluation of posterior according to Bayesian theorem is straightforward, the computation complexity of $p(\ubm{s}| \ubm{x}; \bm{H})$ grows exponentially with the length of $\ubm{s}$. Therefore, we employ forward-backward algorithm \cite{Bishop:2006:PRM:1162264} to do the posterior computation efficiently. As we would detail in next section, it is actually $p(s| \ubm{x}; \bm{H})$ and $p(s, \kappa| \ubm{x}; \bm{H})$ that are needed to formulate the problem. For the joint posterior $p(s, \kappa| \ubm{x}; \bm{H})$, it can be computed by following the Bayesian rule when posterior of hidden state is available.

With such a solution framework ready for model GenHMM, there are still remaining problems to address before it can be employed for practical usage, including:
\begin{itemize}
\item how to realize GenHMM by neural network based generators such that likelihoond of their induced distributions can be computed explicitly and exactly?
\item how to train GenHMM to sovle problem in \eqref{eq:ml-of-hmm} using practical algorithm?
\item would the training of GenHMM converges?
\end{itemize}
We would attempt these problems in the following section.

@@

In this section, we detail the solution for realizing and learning GenHMM. % We begin with introducing flow model that we are going to use to model our generators $\{\bm{g}_{s, \kappa}| s\in\Ss, \kappa=1,2,\cdots, K\}$ for efficient likelihood computation. The learning method of HMM is followed then. 
The convergence analysis is discussed at the end of this section.

\subsection{Realizing $\bm{g}_{s,\kappa}$ by a Flow Model}
We use a feed-forward neural network to implement each generator $\bm{g}_{s,\kappa}$.
Define  $\bm{g}_{s,\kappa}$ as a $L$-layer neural network and we formulate its mapping by each layer's concatenation:
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$, where superscipt $[l]$ denotes the $l$'s layer index and $\circ$ denotes mapping concatenation. Assume $\bm{g}_{s,\kappa}$ is invertible and denotes its inverse mapping as $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$. For a latent variable $\bm{z}$ with density function $p_{s,\kappa}(\bm{z})$, the generated signal $\bm{x}$ follows an induced distribution with density function \eqref{eq:changel-variable}. We illustrate the signal flow between latent variable $\bm{z}$ and observable variable $\bm{x}$ as follows:
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{h}_L=\bm{x}}$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture},
\end{equation*}
where $\bm{f}_{s,\kappa}^{[l]}$ is the $l$-th layer of $\bm{f}_{s,\kappa}$. We have $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. If every layer of $\bm{g}_{s,\kappa}$ is invertible,
the full feed-forward neural network is invertible. Flow model, proposed in \cite{DBLP:journals/corr/DinhKB14} as an image generating model, is such a invertible
feed-forward neural network. It is further improved in subsquential works \cite{2016arXiv160508803D,kingma2018glow} for high-fidelity and high-resolution image generating and representation. Flow model also has advantages as efficient Jacobian computation and low computational complexity. %Computation of inverse mapping for flow-based network and determinant of Jacobin were addressed in the literature.
